Turbo fast no compromise gravitational wave parameter estimation

### 2023-01-21 src/tex/ms.tex
edits up to and inc Sec 3

% Abstract with filler text
\begin{abstract}
We present a light-weight, flexible, and high-performance framework for
inferring the properties of gravitational-wave events. By combining likelihood
heterodyning, automatically-differentiable and accelerator-compatible
waveforms, and gradient-based Markov chain Monte Carlo (MCMC) sampling enhanced
by normalizing flows, we achieve full Bayesian parameter estimation for real
events like GW150914 and GW170817 within a minute of sampling time.  \mi{should
we say a standard analysis takes hours?} Our framework does not require
pre-training or explicit reparameterizations and can be generalized to handle
higher dimensional problems. We present the details of our implementation and
discuss trade-offs and future developments in the context of other proposed
strategies for real-time parameter estimation. Our code for generating the
manuscript and running the analysis is publicly available on GitHub.
\end{abstract}

@@

In this work, we present a light-weight, flexible, and high performance
framework to infer GW event parameters in a fully-Bayesian analysis. Our
framework implements the following major features to achieve its performance:
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item differentiable waveform models,
\item normalizing-flow enhanced Markov chain Monte-Carlo (MCMC) sampler,
\item heterodyned likelihood,
\item native support for hardware accelerators.
\end{enumerate}
The main advantage of our framework is it does not rely on specific models
\mi{what do you mean by ``models'' here? we rely on specific waveforms} or
assumptions about the problem to achieve its performance. This makes our method
extensible to problems beyond the standard CBC analysis, without sacrificing
accuracy for efficiency.
\mi{should we also say something of the advantage of having better, modern
sampling with diffable WFs? can we say we expect this to be more robust to
strange posterior geometries, as you might get by introducing tidal and non-GR
params?}

@@

% Problem with just HMC or MALA
While gradient-based samplers have been shown to outperform gradient-free
algorithms in many practical applications, there remain classes of problems that
most gradient-based samplers do not solve well. For example, first-order
gradient-based algorithms struggle with target distributions that exhibit local
correlations, since they can only handle global structures by construction
\cite{2017arXiv170102434B}.\footnote{Sampling algorithms that use the
information of higher order derivatives such as manifold-MALA and
Riemannian-HMC \cite{RMHMC} can in principle handle local correlations in the
target distribution; however, they often encounter instabilities when used in
real-life applications, so their use is rare practice.} Another example is
multi-modality: if there are multiple modes in the target distribution,
individual chains will likely be trapped in one mode and take an extremely long
time to transverse between the modes \cite{2018arXiv180803230M}.  This means
that the relative weights between modes will take much longer to estimate than
the shape of each mode.

% Long burn-in too
Moreover, before we can use the sampling chain to estimate the posterior
quantities we care about, the sampler often needs to first find the most
probable region in the target space; this is a common process often referred to
as ``burn-in'' in the literature. As a consequence, one would discard a certain
amount of data generated from the beginning of the sampling process, and only
use the later part of the chain to estimate the quantities of interest. The
burn-in phase of a gradient-based sampler is often as long as the sampling
phase, which means that a good portion of the computation is not directly
devoted to estimating the target quantities.

% Crux of normalizing flow
All the above issues can be mitigated by normalizing flows.  Normalizing flows
are a technique based on neural networks that aims at learning a mapping from a
simple distribution, such as a Gaussian, to a complex distribution, often given
in the form of samples \cite{2019arXiv190809257K, 2019arXiv191202762P}. Once
the network is trained, one can evaluate the probability density of the complex
distribution and sample from it very efficiently, by first evaluating the
simple distribution and then applying the learned mapping. The core equation
of normalizing flows is the coordinate transformation of probability
distributions via a Jacobian, as given by
\begin{align}
    p_x(X) = p_z(Z) \left| \frac{\partial f}{\partial z}\right|,
\end{align}
\mi{is there an inverse missing on the Jacobian factor?}
where $p_x(X)$ is the complex target distribution, $p_z(Z)$ is the simple
latent distribution and $f$ is an invertible parameterized transform that
connects the two distributions, $x = f(z)$, to be learned by the normalizing
flow. For a detailed discussion of the algorithm, we refer the readers to
\cite{2019arXiv190809257K, 2019arXiv191202762P}.

% The flowMC algorithm
\mi{I think this following paragraph could be made clearer: what does it mean
to learn the ``global landscape''? and how does this jive with the fact that
norm flows are supposed to help with the \emph{local} correlations?}
Working in tandem, gradient-based MCMC and normalizing flows can efficiently
explore posteriors with local and global correlations, as well as multiple
separate modes.  The scheme relies on iteratively using draws from the
gradient-based MCMC to train a normalizing flow, which is then itself used as a
proposal for another stage of MCMC sampling.

Concretely, we begin by producing initial training data for the normalizing
flow by running multiple independent chains of the gradient-based algorithm for
a fixed number of steps.
From the resulting pool of
samples, the normalizing
flow can begin to learn the global landscape of the target distribution.
%
However, since the independent chains contain the same number of samples, the
relative weight assigned to each chains will differ from the true target distribution (e.g., the relative
importance of separate modes will not be correctly calibrated). This is
mitigated by a second stage of gradient-based MCMC sampling that uses the
distribution learned by the normalizing flow as a \textit{proposal}.

Given a trained normalizing flow model, we can generate the proposed jump in
the target space by sampling from the latent distribution $z \sim p_z(Z)$,
usually a Gaussian, and then pushing it through the learned map given by the
normalizing flow model $x=f(z)$.  The acceptance criterion is then set to be
\begin{align}
    \alpha(\mathbf{x},\mathbf{x_0}) = \min \left[ 1, \frac{\hat{\rho}(\mathbf{x_0})\rho_*(\mathbf{x})}{\hat{\rho}(\mathbf{x})\rho_*(\mathbf{x_0})}\right],
\end{align}
where $\hat{\rho}$ is the probability density estimated by the normalizing flow
model, $\rho_*$ is the probability density evaluated using the target function,
and $x_0$ is the current position.

We can see the flow distribution is the
target distribution when the accepting probability is 1. When the normalizing
flow model has not converged to the target distribution, only a portion of the
proposed jumps will be accepted. This means an MCMC process using the
normalizing flow model as the proposal distribution can adjust the normalization
across different regions of the target parameter space by rejecting jumps into
less likely regions. The training and sampling are then repeated until certain
criteria are met.

@@

\begin{algorithm}
\caption{\textsc{flowMC} pseudocode \mi{can I edit this?}}\label{alg:cap}
\KwIn{initial position $ip$}
\Parameters{n training loop $nt$, n production loop $np$}
\Variables{current chain $cc$, current position $cp$, current NF parameters $\Theta$}
\KwResult{Chains}
$cp \leftarrow ip$\\
\tcc{Training loop}
    \For{$i<nt$}{
        $cc, cp \leftarrow LocalSampling(cp)$\\
        $\Theta \leftarrow TuneNF(cc)$\\
        $chains, cp \leftarrow GlobalSampling(cp, \Theta)$ \\
        $cc \leftarrow Append(cc, chains)$
    }
\tcc{Production loop}
    \For{$i<np$}{
        $c_{local}, cp \leftarrow LocalSampling(cp)$\\
        $c_{global}, cp \leftarrow GlobalSampling(cp, \Theta)$ \\
        $Chains \leftarrow Append(Chains, c_{local}, c_{global})$
    }

\Return{$Chains$}
\end{algorithm}

@@

To demonstrate the robustness of our pipeline, we use it to recover the
parameters of a set of simulated signals. We create a set of simulated signals
and inject them into simulated noise, generated under the Gaussian noise
assumption using the designed PSD for every detector. Then we run our pipeline
given the simulated data and determine the credible interval at the true
parameters of the injected signals. Given the set of credible values, we can
check whether the true value lies within a certain credible interval at a
reasonable frequency. If our pipeline is perfect, we should find the true
parameters lie within $x\%$ credible interval $x\%$ of the time, e.g. the true
value should lie within the $50\%$ credible interval $50\%$ of the time. In
other word, the percentiles of the true parameters should be uniformly
distributed. Deviation from this behavior suggests the pipeline is either
over-confident or too conservative.

% State the distribution of injected population
We sample 1200 events from the following distribution of parameters detailed in
table \ref{tab:parameters}.
The same distribution of parameters are used as the prior in the PE process.
The simulated signals have a 16 seconds segment length,
and we project the signal on the LIGO Hanford, Livingston and Virgo detector.
We used the designed sensitivity curve for each detector to generate the noise \kw{cite}.
We use a minimum frequency cutoff of 30 Hz with a sampling rate of 2048 Hz.

\begin{table*}[hbt!]
    \begin{center}
    \begin{tabular}{ l l l l l }
    \hline
    \hline
    Parameters &  Description & Injection & GW150914 & GW170817\\

    $M_c$ & Chirp mass $[M_\odot]$& $[10, 50]$ & $[10,80]$ & $[1.18,1.21]$ \\
    $q$ & Mass ratio & $[0.5, 1]$ & $[0.125,1]$ & $[0.125,1]$ \\
    $\chi_1$ & Primary dimensionless spin& $[-0.5, 0.5]$ & $[-1,1]$ & $[-0.3,0.3]$ \\
    $\chi_2$ & Secondary dimensionless spin & $[-0.5, 0.5]$ & $[-1,1]$ & $[-0.3,0.3]$ \\
    $d_L$ & Luminosity distance $[\textrm{Mpc}]$ & $[300, 2000]$ & $[0, 2000]^\dag$ & $[1, 75]^\dag$ \\
    $t_c$ & Coalescence time $[\textrm{s}]$& $[-0.5, 0.5]$ & $[-0.1, 0.1]$ & $[-0.1, 0.1]$ \\
    $\phi_c$ & Coalescence phase & $[0, 2\pi]$ & $[0, 2\pi]$ & $[0, 2\pi]$ \\
    $\cos{\iota}$ & Cosine of inclination angle & $[-1, 1]$ & $[-1, 1]$ & $[-1, 1]$ \\
    $\psi$ & Polarization angle & $[0, \pi]$ & $[0, \pi]$ & $[0, \pi]$ \\
    $\alpha$ & Right ascension & $[0, 2\pi]$ & $[0, 2\pi]$ & $[0, 2\pi]$ \\
    $\sin{\delta}$ & Sine of declination & $[-1, 1]$ & $[-1, 1]$ & $[-1, 1]$ \\

    \hline
    \hline
    \end{tabular}
    \caption{Parameters used and the range of prior for the injection-recovery
    test, GW150914, and GW170817. All priors assume a uniform distribution with
    the range tabulated in the table. The coalescence time refers to a shift in
    time around the trigger time. $^\dag$ Instead of uniform in luminosity
    distance, the distance prior used for the GW150914 and GW170817 analysis are
    uniform in comoving distance, with the range denote by the value in the
    table in $\textrm{MPC}$}
    \label{tab:parameters}
    \end{center}
\end{table*}

% Show injection pp plot
The result of this injection-recovery campaign is shown in fig. \ref{fig:ppplot}. We
computed the quantile of which the true value lies for each marginalized
distribution, then plotted the cumulative distribution of these quantiles. The
shadow band denotes the 95\% credible interval drawn from a uniform distribution
with the same number of events. We can see most the injection-recovery result
lies within the band drawn from a uniform distribution, showing our inference
results agree well with a uniform distribution. There is a small deviation from
a uniform distribution for the secondary spin $\chi_2$, which is not too
alarming since we are computing the quantile for 11 parameters. Having one of
them lying briefly outside the 95\% band does not mean the inference is bias.

To quantify how well our result agrees with a uniform distribution, we can
compute the Kolmogorov-Smirnov test p-values for each marginalized distribution.
If the p-value is too low, often chosen to be $0.05$, then it means our result
does not agree with a uniform distribution, hence it could be biased. The
p-values for each marginalized distribution are shown in the legend of fig.
\ref{fig:ppplot}. We can see most of them are well above the $0.05$ threshold,
except for $\chi_2$, which is mildly below the threshold. Once again, assuming
these p-values are drawn from a uniform distribution, given $11$ draws (the
number of parameters in our inference), it is not abnormal to have one of the
parameters lies slightly outside the threshold. To assess whether this is
expected, we can compute the combined p-value for these 11 parameters, which is $0.47$.
This shows our inference pipeline performs properly on simulated data.
Indeed, the accuracy of our inference pipeline is consistent with other works\kw{cite}.

@@

To demonstrate the performance of our parameter estimation pipeline, we apply
our pipeline to analyze GW150914 and GW170817. The prior used for the two events
are shown in table \ref{tab:parameters}. We use a 4 seconds long segment for the
GW150914 analysis, and a 128 seconds long segment for the GW170817 analysis.
Both data segment are fetched from GWOSC \kw{cite}. Given the specific sampler
setting we use, we produce around an average of 2500 and 3500 \textbf{effective
samples}\footnote{Note that effective sample measures the number of independent
samples, which takes correlation between step into account, and it is not the
same as the number of generated samples. The effective sample size is computed
using
\texttt{arviz},\url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}.}
for GW150914 and GW170817 in each analysis respectively, which is run on an
Nvidia A100 GPU. The \texttt{wall} time for both event is around 10 minutes. The
chain data and the analysis scripts which generate the chains can be found on
%\kw{\variable{data/} and \variable{script}}
. Most of the time is spent on jit compilation of the code, and the actual
sampling time was about \~150s. Note that we have pre-computed the reference
waveform parameters used in hetetrodyne likelihood for the two events, so the
time spent on solving for the reference waveform parameters is omitted in the
\texttt{wall} time calculation.

\begin{figure}
\includegraphics[width=0.99\linewidth]{static/GW150914.png}
\caption{
    hi
}
\label{fig:GW150914}
\end{figure}

\begin{figure}
\includegraphics[width=0.99\linewidth]{static/GW170817.png}
\caption{
    hi
}
\label{fig:GW170817}
\end{figure}

Since we are using the \texttt{IMRPhenomD} waveform and the open data release
available on GWOSC \kw{cite} does not provide the posterior samples for this
waveform, we run Bilby using the same prior and waveform model as a reference
for comparison. From fig \ref{fig:GW150914} and \ref{fig:GW170817}, we can see
the 11-dimensional posteriors results produced by our pipeline are consistent
with the results obtained using Bilby \kw{cite}.

@@

For a quantitative comparison between results obtained through our code and
other existing tools, we compute the Jensen-Shannon divergence of the
marginalized distribution between our code and \texttt{Bilby}. The
Jensen-Shannon divergence is a symmetric measure of the distance between two
probability distributions, with a value of 0 indicating identical distributions
and a value of $\ln{2} \textrm{nat}$ representing the maximum possible
divergence between two distributions.

@@

% Brief summary
In this work, we present a PE pipeline for GW events that is efficient and
flexible. Our package put together a number of innovations, including
differentiable waveform models from \texttt{ripple} \kw{cite}, heterodyned
likelihood and normalizing flow enhanced sampler \texttt{flowMC}. We tested the
robustness of our pipeline on a set of 1200 synthetic GW events, showing it is
capable of handling catalog that will be available in the near future. We also
demonstrate our pipeline can estimate the parameters of GW150914 and GW170817 in
\kw{Quote speed}.

% Compared to other works
There are recent works from different groups on speeding up parameter estimation
of gravitational wave, including efficient reparameterization
\cite{Islam:2022afg,Roulet:2022kot} and deep learning
\cite{Dax:2021tsq,Dax:2022pxd}. While all of these methods can get down to
minutes-scale parameter estimation with high fidelity, we would like to
highlight the unique strength of this work, and the potential interplay between
our work and others' work.

@@

Another overhead we are looking into reducing is the time needed to find the
reference waveform for heterodyning the likelihood. Currently, we are using
\texttt{differential evolution} \cite{Storn1997DifferentialE} available in
\texttt{scipy} to find the waveform parameters which maximize the likelihood.
Since \texttt{differnetial evolution} has not been implemented in Jax, and the
Jax waveform we use is not compatible with the parallelization scheme in the
\texttt{scipy} library, it takes around 5 minutes \kw{Benchmark it in the code
and make this number precise.} to find the reference waveform parameters for
GW170817. There are two ways to improve the performance in finding the
parameters of the reference waveform. First, we can explore different
optimization strategy that is more compatible with the strength of our pipeline,
in particular differentiability of our likelihood and the possibility to
evaluate many waveforms in parallel with a GPU. Particle swarm \cite{7869491}
and stochastic gradient descent methods \cite{10.5555/304710.304720} are
promising candidates which we intend to investigate in the future.

>>>>

% Abstract with filler text
\begin{abstract}
We present a lightweight, flexible, and high-performance framework for
inferring the properties of gravitational-wave events. By combining likelihood
heterodyning, automatically-differentiable and accelerator-compatible
waveforms, and gradient-based Markov chain Monte Carlo (MCMC) sampling enhanced
by normalizing flows, we achieve full Bayesian parameter estimation for real
events like GW150914 and GW170817 within a minute of sampling time.  \mi{should
we say a standard analysis takes hours?} Our framework does not require
pre-training or explicit reparameterizations and can be generalized to handle
higher dimensional problems. We present the details of our implementation and
discuss trade-offs and future developments in the context of other proposed
strategies for real-time parameter estimation. Our code for generating the
manuscript and running the analysis is publicly available on GitHub.
\end{abstract}

@@

In this work, we present a lightweight, flexible, and high performance
framework to infer GW event parameters in a fully-Bayesian analysis. Our
framework implements the following major features to achieve its performance:
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item differentiable waveform models,
\item normalizing-flow enhanced Markov chain Monte-Carlo (MCMC) sampler,
\item heterodyned likelihood,
\item native support for hardware accelerators.
\end{enumerate}
The main advantage of our framework is it does not rely on specific models
\mi{what do you mean by ``models'' here? we rely on specific waveforms} or
assumptions about the problem to achieve its performance. This makes our method
extensible to problems beyond the standard CBC analysis, without sacrificing
accuracy for efficiency.
\mi{should we also say something of the advantage of having better, modern
sampling with diffable WFs? can we say we expect this to be more robust to
strange posterior geometries, as you might get by introducing tidal and non-GR
params?}

@@

% Problem with just HMC or MALA
While gradient-based samplers have been shown to outperform gradient-free
algorithms in many practical applications, there remain classes of problems that
most gradient-based samplers do not solve well. For example, first-order
gradient-based algorithms struggle with target distributions that exhibit locally-varying
correlations, since they assume a uniform global structure by construction
\cite{2017arXiv170102434B}.\footnote{Sampling algorithms that use the
information of higher order derivatives such as manifold-MALA and
Riemannian-HMC \cite{RMHMC} can in principle handle local correlations in the
target distribution; however, they often encounter instabilities when used in
real-life applications, so their use is rare practice.} Another example is
multi-modality: if there are multiple modes in the target distribution,
individual chains will likely be trapped in one mode and take an extremely long
time to transverse between the modes \cite{2018arXiv180803230M}.  This means
that the relative weights between modes will take much longer to estimate than
the shape of each mode.

% Long burn-in too
Moreover, before we can use the sampling chain to estimate the posterior
quantities we care about, the sampler often needs to first find the most
probable region in the target space (known as the \emph{typical set}); this is
a common process often referred to as ``burn-in'' in the literature. As a
consequence, one would discard a certain amount of data generated from the
beginning of the sampling process, and only use the later part of the chain to
estimate the quantities of interest. The burn-in phase of a gradient-based
sampler is often as long as the sampling phase, which means that a good portion
of the computation is not directly devoted to estimating the target quantities.

% Crux of normalizing flow
All the above issues can be mitigated by normalizing flows.  Normalizing flows
are a technique based on neural networks that aims at learning a mapping from a
simple distribution, such as a Gaussian, to a complex distribution, often given
in the form of samples \cite{2019arXiv190809257K, 2019arXiv191202762P}. Once
the network is trained, one can evaluate the probability density of the complex
distribution and sample from it very efficiently, by first evaluating the
simple distribution and then applying the learned mapping. The core equation
of normalizing flows is the coordinate transformation of probability
distributions via a Jacobian, as given by
\begin{align}
    p_x(X) = p_z(Z) \left| \frac{\partial f}{\partial z}\right|^{-1},
\end{align}
where $p_x(X)$ is the complex target distribution, $p_z(Z)$ is the simple
latent distribution and $f$ is an invertible parameterized transform that
connects the two distributions, $x = f(z)$, to be learned by the normalizing
flow. For a detailed discussion of the algorithm, we refer the readers to
\cite{2019arXiv190809257K, 2019arXiv191202762P}.

% The flowMC algorithm
\mi{can we clarify what it means to learn the ``global landscape'', and how
this jives with the fact that norm flows are supposed to help with the
\emph{local} correlations?}
Working in tandem, gradient-based MCMC and normalizing flows can efficiently
explore posteriors with local and global correlations, as well as multiple
separate modes.  The scheme relies on iteratively using draws from the
gradient-based MCMC to train a normalizing flow, which is then itself used as a
proposal for another stage of MCMC sampling.

Concretely, we begin by producing initial training data for the normalizing
flow by running multiple independent chains of the gradient-based algorithm for
a fixed number of steps.  From the resulting pool of samples, the normalizing
flow can begin to learn the global landscape of the target distribution.
%
However, since the independent chains contain the same number of samples, the
relative weight assigned to each chains will not represent the true target
distribution (e.g., the relative importance of separate modes will not be
correctly calibrated). This is mitigated by a second stage of gradient-based
MCMC sampling that uses the distribution learned by the normalizing flow as a
\textit{proposal}.

Given a trained normalizing flow model, we can generate the proposed jump in
the target space by sampling from the latent distribution $z \sim p_z(Z)$,
usually a Gaussian, and then pushing it through the learned map given by the
normalizing flow model $x=f(z)$.  The acceptance criterion is then set to be
\begin{align} \label{eq:flow-proposal}
    \alpha(\mathbf{x},\mathbf{x_0}) = \min \left[ 1, \frac{\hat{\rho}(\mathbf{x_0})\rho_*(\mathbf{x})}{\hat{\rho}(\mathbf{x})\rho_*(\mathbf{x_0})}\right],
\end{align}
where $\hat{\rho}$ is the probability density estimated by the normalizing flow
model, $\rho_*$ is the probability density evaluated using the target function,
and $x_0$ is the current position.

From Eq.~\eqref{eq:flow-proposal}, we can see that the flow distribution is the
target distribution when the accepting probability is 1. When the normalizing
flow model has not converged to the target distribution, only a portion of the
proposed jumps will be accepted. This means an MCMC process using the
normalizing flow model as the proposal distribution can adjust the
normalization across different regions of the target parameter space by
rejecting jumps into less likely regions. The training and sampling are then
repeated until certain criteria, at each step combining global and local MCMC
sampling which respectively do and do not use the normalizing flow as proposal.

@@

\begin{algorithm}
\caption{\textsc{flowMC} pseudocode}\label{alg:cap}
\KwIn{initial position $ip$}
\Parameters{number of training loops $nt$, number of production loops $np$}
\Variables{current chain $cc$, current position $cp$, current NF parameters $\Theta$}
\KwResult{$chains$}
$cp \leftarrow ip$\\
\tcc{Training loop}
    \For{$i<nt$}{
        $cc, cp \leftarrow LocalSampling(cp)$\\
        $\Theta \leftarrow TuneNF(cc)$\\
        $chains, cp \leftarrow GlobalSampling(cp, \Theta)$ \\
        $cc \leftarrow Append(cc, chains)$
    }
\tcc{Production loop}
    \For{$i<np$}{
        $c_{local}, cp \leftarrow LocalSampling(cp)$\\
        $c_{global}, cp \leftarrow GlobalSampling(cp, \Theta)$ \\
        $chains \leftarrow Append(chains, c_{local}, c_{global})$
    }

\Return{$chains$}
\end{algorithm}

@@

To demonstrate the robustness of our pipeline, we use it to recover the
parameters of a set of simulated signals. We create a set of simulated signals
and inject them into simulated stationary Gaussian noise.
Then we run our pipeline on the simulated data, and determine the credible
interval at which the true parameters of the injected signals are recovered. From the set of
credible values, we can check whether the true value lies within a certain
credible interval at the expected frequency: if our pipeline is working as expected, we
should find the true parameters lie within $x\%$ credible interval $x\%$ of the
time, e.g., the true value should lie within the $50\%$ credible interval $50\%$
of the time. In other word, the percentiles of the true parameters should be
uniformly distributed. Deviation from this behavior suggests the pipeline is
either over-confident or too conservative \cite{Cook2006,Talts2018}.

% State the distribution of injected population
We sample 1200 events from the distribution of parameters detailed in Table
\ref{tab:parameters}; the same distributions are used as the prior in the PE
process.  We simulate signals over 16 seconds of data, with a minimum frequency
cutoff of 30 Hz and a sampling rate of 2048 Hz.  We draw noise from the design
PSDs for the LIGO Hanford, LIGO Livingston and Virgo detectors \cite{} \mi{what
curves did you use?}.
For both injection and recovery, we make use of the \textsc{IMRPhenomD}
waveform \cite{Khan:2015jqa} via the fully-differentiable implementation
presented in the \textsc{ripple} package \cite{}.

\begin{table*}[hbt!]
    \begin{center}
    \begin{tabular}{ l l l l l }
    \hline
    \hline
    Parameter &  Description & Injection & GW150914 & GW170817\\
    \hline

    $M_c$ & chirp mass $[M_\odot]$& $[10, 50]$ & $[10,80]$ & $[1.18,1.21]$ \\
    $q$ & mass ratio & $[0.5, 1]$ & $[0.125,1]$ & $[0.125,1]$ \\
    $\chi_1$ & primary dimensionless spin& $[-0.5, 0.5]$ & $[-1,1]$ & $[-0.3,0.3]$ \\
    $\chi_2$ & secondary dimensionless spin & $[-0.5, 0.5]$ & $[-1,1]$ & $[-0.3,0.3]$ \\
    $d_L$ & luminosity distance $[\textrm{Mpc}]$ & $[300, 2000]$ & $[0, 2000]^\dag$ & $[1, 75]^\dag$ \\
    $t_c$ & coalescence time $[\textrm{s}]$& $[-0.5, 0.5]$ & $[-0.1, 0.1]$ & $[-0.1, 0.1]$ \\
    $\phi_c$ & coalescence phase & $[0, 2\pi]$ & $[0, 2\pi]$ & $[0, 2\pi]$ \\
    $\cos{\iota}$ & cosine of inclination angle & $[-1, 1]$ & $[-1, 1]$ & $[-1, 1]$ \\
    $\psi$ & polarization angle & $[0, \pi]$ & $[0, \pi]$ & $[0, \pi]$ \\
    $\alpha$ & right ascension & $[0, 2\pi]$ & $[0, 2\pi]$ & $[0, 2\pi]$ \\
    $\sin{\delta}$ & sine of declination & $[-1, 1]$ & $[-1, 1]$ & $[-1, 1]$ \\

    \hline
    \hline
    \end{tabular}
    \caption{Prior ranges for parameters varied in the injection-recovery
    test, as well as the GW150914 and GW170817 analyses. All priors are uniform over the ranges shown, except for the luminosity distance prior in the GW150914 and GW170817 analyses ($^\dag$) for which we apply a prior unform in comoving volume. The coalescence time refers to a shift relative to the geocenter trigger time, and $M_c$ refers to the redshifted (detector-frame) chirp mass.}
    \label{tab:parameters}
    \end{center}
\end{table*}

% Show injection pp plot
We summarize the result of this injection-recovery campaign in
Fig.~\ref{fig:ppplot}. This shows the cumulative distribution over injections
of the quantile at which the true value lies in the marginalized distribution
of each parameter. The shaded band denotes the 95\%-confident variation
expected from draws from a uniform distribution with the same number of events.
We can see that most of the measured curves lie within this band, showing that
our inference results agree well with a uniform distribution. There is a small
deviation from a uniform distribution for the secondary spin $\chi_2$, which is
not alarming given that we are computing the quantile for 11 parameters.  This
is what we expect if our pipeline is working as expected.

To further quantify how well our result agrees with a uniform distribution, we
can compute the Kolmogorov-Smirnov $p$-values for each marginalized
distribution \cite{}.  If this $p$-value were is low (with a threshold often
chosen to be $p = 0.05$), then our result could be in tension with a uniform
distribution. The $p$-values obtained for each parameter are shown in the
legend of Fig.~\ref{fig:ppplot}.  Most of them are well above the $p = 0.05$
threshold, except for $\chi_2$, which is mildly below the threshold. Once
again, assuming these $p$-values are drawn from a uniform distribution, given
11 draws (the number of parameters in our inference), it is not abnormal to
have one of the parameters lying slightly outside the threshold. To assess
whether this is expected, we can compute the combined $p$-value for these 11
parameters, and find it to be $p = 0.47$.  This shows our inference pipeline
performs properly on simulated data at a similar level as standard tools
\cite{Veitch:2014wba,Romero-Shaw:2020owr}.

@@

To demonstrate the performance of our parameter estimation pipeline, we apply
it to two real LIGO-Virgo events: GW150914 and GW170817. We use the priors
shown in Table \ref{tab:parameters}, and take 4 s of data sampled at 2048 Hz
for the GW150914 analysis, and 128 s of data sampled at 4096 Hz for the
GW170817 analysis; strain data and PSDs for both events are fetched from GWOSC
\cite{GWOSC}. For our specific choice of sampler settings, we produce
${\sim}$2500 and 3500 \emph{effective samples}\footnote{Effective
samples here refers to the number of independent samples, which is the total
number of generated samples divided by their correlation length; we compute the
effective sample size using
\textsc{arviz},\url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}.
\mi{cite}} for GW150914 and GW170817 respectively. Running on an Nvidia A100
GPU, the wall time for both events is around 10 minutes. Most of this
time is spent on just-in-time (JIT) compilation of the code; the actual
sampling time is only ${\sim}150$ s. We pre-compute the reference waveform
parameters used in hetetrodyne likelihood for the two events, the time for
which is omitted in the \texttt{wall} time calculation.  The chain data and the
analysis scripts which generate the chains can be found in
%\kw{\variable{data/} and \variable{script}}
.

\begin{figure}
\includegraphics[width=0.99\linewidth]{static/GW150914.png}
\caption{
    GW150914 posterior computed by our code (\mi{blue}) and \textsc{Bilby} (\mi{orange}).
}
\label{fig:GW150914}
\end{figure}

\begin{figure}
\includegraphics[width=0.99\linewidth]{static/GW170817.png}
\caption{
    GW170817 posterior computed by our code (\mi{blue}) and \textsc{Bilby} (\mi{orange}).
}
\label{fig:GW170817}
\end{figure}

For comparison, we produce equivalent runs with \textsc{Bilby}, using the same
exact data and priors.  We use the \textsc{dynesty} sampler
\cite{2020MNRAS.493.3132S,dynesty}, with 1000 live points and other settings as
shown in \cite{release}.  We carry out these runs using \textsc{parallel Bilby}
(\textsc{pBilby}) \cite{Smith:2019ucc} to distribute the computation over 400
Intel Skylake CPUs for each event.  For the specific settings chosen, the
wall-time duration of each run is ${\sim}1$ day for GW150914 and ${\sim}2$ days
for GW170817.

@@

Figs.~\ref{fig:GW150914} and \ref{fig:GW170817} show that our posteriors are
consistent with those produced by \textsc{Bilby}.  For a quantitative
comparison, we compute the Jensen-Shannon divergence between our code and
\textsc{Bilby} for the marginalized distribution for each parameter. The
Jensen-Shannon divergence is a symmetric measure of the distance between two
probability distributions, with a value of 0 indicating identical distributions
and a value of $\ln{2} \textrm{nat}$ representing the maximum possible
divergence between two distributions.

@@

% Brief summary
In this work, we present a PE pipeline for GW events that is efficient and
flexible. Our package put together a number of innovations, including
differentiable waveform models from \texttt{ripple} \kw{cite}, heterodyned
likelihood and normalizing flow enhanced sampler \texttt{flowMC}. We tested the
robustness of our pipeline on a set of 1200 synthetic GW events, showing it is
capable of handling catalogs that will be available in the near future. We also
demonstrate our pipeline can estimate the parameters of GW150914 and GW170817 in
\kw{Quote speed}.

% Compared to other works
There are recent works from different groups on speeding up parameter
estimation of gravitational wave, including efficient reparameterization
\cite{Islam:2022afg,Roulet:2022kot} and deep learning
\cite{Dax:2021tsq,Dax:2022pxd}. While all of these methods can get down to
minutes-scale parameter estimation with high fidelity, our approach possesses
unique strengths, and may complement some of those other techniques.

@@

Another overhead we are looking into reducing is the time needed to find the
reference waveform for heterodyning the likelihood. Currently, we are using
\texttt{differential evolution} \cite{Storn1997DifferentialE} available in
\texttt{scipy} to find the waveform parameters which maximize the likelihood.
\mi{we should say this in the het like sect}
Since \texttt{differnetial evolution} has not been implemented in Jax, and the
Jax waveform we use is not compatible with the parallelization scheme in the
\texttt{scipy} library, it takes around 5 minutes \kw{Benchmark it in the code
and make this number precise.} to find the reference waveform parameters for
GW170817. There are two ways to improve the performance in finding the
parameters of the reference waveform. First, we can explore different
optimization strategy that is more compatible with the strength of our pipeline,
in particular differentiability of our likelihood and the possibility to
evaluate many waveforms in parallel with a GPU. Particle swarm \cite{7869491}
and stochastic gradient descent methods \cite{10.5555/304710.304720} are
promising candidates which we intend to investigate in the future.
